---
title: "Introduction to Bayesian Computation Using the rstanarm R Package"
author: Ben Goodrich of Stan Group (http://stan.fit)
date: "November 22, 2016"
autosize: true
output: 
  ioslides_presentation:
    widescreen: true
---

```{r, setup, include = FALSE}
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(png))
suppressPackageStartupMessages(library(grid))

opts_chunk$set(dev.args = list(pointsize = 18), 
               warning = FALSE, message = TRUE)
```

## Obligatory Disclosure

* Ben is an employee of Columbia University, which has received several research grants to develop Stan
* Ben is also a cofounder of Stan Group (http://stan.fit), which provides support, consulting, etc. for 
  businesses using Stan
* According to Columbia University policy, any such employee who has any equity stake in, a title (such as 
  officer or director) with, or is expected to earn at least $\$5,000.00$ per year from a 
  private company is required to disclose these facts in presentations

## Installation of the rstanarm R Package

* Try simply executing
```{r, eval = FALSE}
install.packages("rstanarm", repos = "https://cloud.r-project.org", dependencies = TRUE)
```
* It may ask you if you want to install a newer version (2.13.1) of rstanarm "from source", in which
  case you should press n (for no) unless you are quite certain that you have a C++ compiler installed
  and configured correctly

## What is Stan?

* Includes a high-level
  [probabilistic programming language](https://en.wikipedia.org/wiki/Probabilistic_programming_language)
  that overlaps a lot with https://cran.r-project.org/view=Distributions
* Includes a translator of high-level Stan syntax to somewhat low-level C++
* Includes a matrix and scalar math library that supports autodifferentiation
* Includes new (and old) gradient-based algorithms for statistical inference
* Includes interfaces from R, such as the **rstanarm** package, and other high-level software
* Includes (not Stan specific) post-estimation R functions
* Includes a large community of users and many developers

## What Is the **rstanarm** R Package

* An R interface to a handful of pre-written, pre-compiled Stan programs
* Syntax mirrors that of popular model-fitting functions in R, including 
    * `lm` and `aov`
    * `glm` and `MASS::glm.nb`
    * `MASS::polr`
    * `lme4::lmer` and `lme4::glmer`
    * `gamm4::gamm4`
* You can use Stan without knowing the Stan language
* Minimal installation requirements; i.e. no C++ compiler

# Why Should You Use Stan for Bayesian Inference?

## Basics of Bayesian Decision Theory

* Quantitative estimation in academia and business / policy differs 
    * In complexity but more importantly 
    * Academics rarely use the estimates to make a potentially costly decision
* Business have to make decisions to maximize _expected_ profit, netting out costs and revenues
* Taking expectations averages over uncertainty in the unknowns

## The Only Four Sources of Uncertainty

1. __Model uncertainty__: Which model(s) should be used?
2. __Parameter uncertainty__: What are the parameter values in the model(s)?
3. __Systematic uncertainty__: What system of reasoning should be used to evaluate the implications of the
  good model(s)?
4. __Software uncertainty__: Does the software successfully estimate the parameters?

* Bayesian approach accounts for the first 3. Probability is the system of reasoning so both model and
  parameter uncertainty are expressed in terms of probability. Stan substantially mitigates (4).
* Supervised learning approaches leave (3) open, typically ignore (2), focus on the _choice_ of the best 
  model but do not quantify model uncertainty
* Frequentist approach uses probability for (3) but interprets it over repeated applications of the 
  estimator to independent samples from the same population. So, it does not really quantify model or 
  parameter uncertainty

## The Only Four Sources of Uncertainty

* To update your beliefs using Bayes' rule you need four things:
    1. A probability distribution for the data-generating process that typically depends on parameters
    2. A description of your current beliefs about those parameters using probability distributions
    3. Data, although it need not be a random sample from a well-defined population
    4. An algorithm to draw from the implied posterior distribution
    
* Stan does (4) very well for many differentiable posterior distributions but you are on your own for 1 -- 3
* The **rstanarm** R package provides a menu of choices for 1 and 2

## Baysian Workflow
```{r fig.width=10, fig.height=5.3, echo=FALSE}
img <- readPNG("bayesian_workflow.png")
 grid.raster(img)
```

# Example with data from LendingClub.com

## LendingClub.com Data

```{r}
source("get_data.R", verbose = FALSE)
Loans <- get_loans_data()
dim(Loans)
training <- Loans[1:10000, ]
testing  <- Loans[-c(1:10000), ]
table(training$y)
```

## Continuous Predictors

```{r}
summary(training[,c("loan_amnt", "annual_inc")])
```

## Binary Predictors

```{r}
with(training, table(term, home_ownership, emp_length != 0))
```

## Loading the **rstanarm** R Package

```{r}
library(rstanarm)
options(mc.cores = parallel::detectCores())
```


## Fitting to Simulated Data

```{r}
y_tilde <- with(training, rbinom(nrow(training), size = 1, 
                                prob = ifelse(emp_length == 0, 0.5, 0.1)))
table(y_tilde)
```

```{r, eval = FALSE}
post_sim <- stan_glm(y_tilde ~ I(training$emp_length != 0), 
                     family = binomial(link = "logit"),
                     prior = normal(), prior_intercept = normal())
```

## A Richer Model for Nonrepayment

```{r, GLM, results='hide', cache=TRUE}
post <- stan_glm(y ~ log(loan_amnt) + term + home_ownership +
                   log(annual_inc) + I(emp_length != 0), 
                 family = binomial(link = "logit"), data = training,
                 prior = normal(), prior_intercept = normal(), QR = TRUE)
```

## Model Graphical Output

```{r, message = FALSE}
include <- c("log(loan_amnt)", "term 60 months", "home_ownershipOWN",
              "home_ownershipRENT", "log(annual_inc)", "I(emp_length != 0)TRUE")
plot(post, prob = 0.80, pars = include)
```

## Using Shinystan

```{r, eval = FALSE}
launch_shinystan(post)
```

## Update Your Beliefs about Residence Variables

```{r, PVP2, message = FALSE, cache=TRUE}
posterior_vs_prior(post, prob = 0.50, regex_pars = "^[h]")
```

## Calculating the Distribution of Profit

* Assume for simplicity (and conservatism) that if a loan is defaulted on, 
  lender gets no interest & loses all principal

```{r}
ppd <- posterior_predict(post, newdata = testing, draws = 500)
profit <- sapply(1:ncol(ppd), FUN = function(i) {
  ifelse(ppd[,i] == 1, -testing$loan_amnt[i], testing$installment[i] *
           ifelse(testing$term[i] == "36 months", 36, 60) -
         testing$loan_amnt[i])
})
```

## Plotting the Distribution of Profit Margin

```{r}
par(mar = c(5,4,1,1), las = 1)
hist(rowSums(profit) / sum(testing$loan_amnt), prob = TRUE, main = "Testing Dataset", 
     xlab = "Profit Margin")
```

## Model Comparison with the loo Package {.smaller}

```{r, comparison, cached = TRUE}
post2 <- update(post, formula. = . ~ . - I(emp_length != 0))
loo(post)
loo(post2)
```

## Wrapup

* Learned a bit about Bayesian theory
* Estimated a relevant model using the **rstanarm** package
* Looked at plots of the model output
* Looked at functions of model output for business decisions
* The **rstanarm** package provides a few models; almost any model can be estimated
  with Bayesian methos if you write it in the Stan language

## Where to Start / Get Help?

* Textbook: http://xcelab.net/rm/statistical-rethinking/
* Forthcoming (in 2017) second edition of textbook using __rstanarm__: http://www.stat.columbia.edu/~gelman/arm/
* User manual: http://mc-stan.org/documentation/
* Examples of Stan programs: https://github.com/stan-dev/example-models
* Mailing list: https://groups.google.com/forum/#!forum/stan-users
* Stan conference: http://mc-stan.org/events/stancon.html
* Commercial support / training: askus@stan.fit
